Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support).

### 2017-02-10 csf-2017/paper.tex
Finish the discussion of the stack and return pointer management stuff.

The basic idea of our approach is quite simple: we stick to a single, rather
standard C stack and register-passed return pointers in much the same way as a
standard C ABI. However, to prevent various ways of misusing this basic scheme,
we put local capabilities to work and take a number of not-always-obvious safety
measures. In the next paragraphs, we will explain the issues to be considered in
all the relevant situations: when (1) starting our program, (2) invoking the
adversary, (3) returning from the adversary, (4) invoking an adversary callback
and (5) having a callback invoked by the adversary.

@@

\textbf{Invoking the adversary}
During execution, we use the stack as normal: we may push values onto it and pop
them back off (e.g. for register spilling etc.). When we invoke the adversary,
we need to take great care to only provide him with the authority we want him to
have. Specifically, we clear all the registers, except those that contain
arguments the adversary is intended to have access to. We leave a copy of the
stack pointer in the appropriate register, but only after we have used the
$\subseg{}{}{}$ to shrink its authority to just the part of the stack that we
are not using. In one of the registers, we will also provide a return pointer.
We will go into constructing the return pointer further below.

It is already important to note that the return pointer
will be provided to the adversary as local capabilities. If the adversary were
able to store away the return pointer in a global data structure, then he would
be able to jump to it in circumstances where this should not be possible: he
may, for example, store the return pointer, legally jump to it a first time,
wait to be invoked again and then jump to the old return pointer a second time,
rather than the new return pointer received for the new invocation. Similarly,
he may store the return pointer, invoke a function in the trusted code, wait to
be invoked again and then jump to the stored return pointer rather than the new
return pointer for the second invocation.

Because the return pointer is local, the adversary is only able to store it
through write-local capabilities, which (because of our assumptions above)
means: on the stack. However, this makes it crucial that the stack pointer is
also local: if the adversary were able to store the stack pointer in a global
data structure, he/she can store return pointers on the stack, and play similar
mischief as mentioned above. Storing away stack pointers for use during future
invocations would also be dangerous for other reasons: during such a future
invocation, a larger part of the stack may be in use by the trusted code, which
the adversary may still have access to through the old stack pointer.

A final measure we need to take is that before invocation of the adversary, we
need to clear (i.e. overwrite with zeros) the part of the stack that we are not
using.  Otherwise, that part of the stack may still contain data and
capabilities from previously executed code, including local capabilities since
they are allowed on the stack.  The adversary would be able to take advantage of
this by using it to store either stack pointers or return pointers from previous
invocations and play the same tricks as above.

\textbf{Returning from the adversary}
So return pointers must be passed as local capabilities. But what should their
permissions be, what memory should they point to and what should that memory
(let's call it the activation record) contain? Let us answer the last question
first by considering what should happen when the adversary jumps to a return
pointer and what capabilities we need for achieving that. Essentially, the
program counter should be restored to the code after the call into the
adversary, so the activation record should store this program counter.
Additionally, the stack pointer should also be restored to its original value.
Since the adversary has a more restricted authority over the stack than the code
making the call, we cannot hope to reconstruct the return pointer from the stack
pointer owned by the adversary. Instead, the old stack pointer should also be
stored as part of the activation record.

@@

It's important that neither of these capabilities are accessible to the
adversary. In other words, we need to provide the adversary a capability that he
can jump to but not read from, i.e. an enter capability. To make this work, we
construct the activation record as depicted in
Figure~\ref{fig:activ-rec-struct}. The return pointer is an $\entry$ capability
pointing with authority over a data structure containing the previous return and
stack pointer as well as a number of restore instructions. The current address
points to those instructions, so that, upon invocation, they can load the old
stack pointer and program counter into the register file through the authority
of the pc. The return pointer is given to the adversary as an enter pointer so
that the adversary cannot get hold of the contents of the data structure, but
after invocation, its permission is updated to $\exec$, so that the activation
record's contents become available to the restore instructions.

The final question that remains is: where should we construct this return
pointer? The attentive reader will see that there is only one possibility: since
the activation record contains the old stack pointer, which is necessarily
local, the activation record can only be constructed in a part of memory where
we have write-local access, i.e. on the stack. Note that this means we will be
placing and executing instructions on the stack, i.e. it will not just contain
code pointers and data.  This means that this discipline should be combined with
protection against stack smashing attacks (i.e. buffer overflows on the stack overwriting activation
records' contents), but capability machines' fine-grained memory protection
luckily is well-suited for this.

\textbf{Invoking an adversary callback}
* must be global

\paragraph{Having a callback invoked by the adversary}
* checking the received stack pointer

\begin{itemize}
\item Issues to consider:
\begin{itemize}
\item no global store-local capabilities
\item clearing the stack
\item clearing (some) registers
\item constructing return pointers
\item constructing callbacks
\item invoking the adversary's callbacks
\end{itemize}
\item Reusable macro-instructions for stack and return pointer management
\begin{itemize}
\item taking care of the stuff above
\end{itemize}
\end{itemize}

>>>>

The basic idea of our approach is quite simple: we stick to a single, rather
standard C stack and register-passed return pointers in much the same way as a
standard C ABI. However, to prevent various ways of misusing this basic scheme,
we put local capabilities to work and take a number of not-always-obvious safety
measures. In the next paragraphs, we will explain the issues to be considered in
all the relevant situations: when (1) starting our program, (2) Returning to the
adversary, (3) invoking the
adversary, (4) returning from the adversary, (5) invoking an adversary callback
and (6) having a callback invoked by the adversary.

@@

\textbf{Returning to the adversary} Perhaps the simplest scenario is the case of
returning to the adversary after he/she invoked our code. In this case, we will
have received a return pointer from him, and we just need to jump to it as
usual. The only thing we need to take care of is to properly clear the registers
before we do (erasing any data we do not want to pass as (part of) the return
value) and similarly, erase any part of the stack that we have used (i.e.
overwrite it with zeros). Note that we can use the stack as normal: we may push
values onto it and pop them back off (e.g. for register spilling etc.).

\textbf{Invoking the adversary} Slightly more complex is the case where we
invoke the adversary. As above, we clear all the registers, except those that
contain arguments the adversary is intended to have access to. We leave a copy
of the stack pointer in the appropriate register, but only after we have used
the $\subseg{}{}{}$ to shrink its authority to just the part of the stack that
we are not using. In one of the registers, we also provide a return pointer,
which we will say more about below.

Already important is the fact that the return pointer must be provided to the
adversary as a local capability. If not, the adversary would be able to store
away the return pointer in a global data structure, and he would be able to jump
to it in circumstances where this should not be possible: he may, for example,
store the return pointer, legally jump to it a first time, wait to be invoked
again and then jump to the old return pointer a second time, rather than the new
return pointer received for the new invocation. Similarly, he may store the
return pointer, invoke a function in our code, wait for us to invoke him again
and then jump to the stored, old return pointer rather than the new one he
received upon the second invocation.

Because the return pointer is local, this is not possible: the adversary can
only store it through write-local capabilities, which (because of our
assumptions above) means: on the stack. However, this makes it crucial that the
stack pointer is also local: if the adversary were able to store the stack
pointer in a global data structure, he/she can store return pointers on the
stack, and play similar mischief as mentioned above. Storing away stack pointers
for use during future invocations would also be dangerous for other reasons:
during such a future invocation, a larger part of the stack may be in use by the
trusted code, which the adversary may still have access to through the old stack
pointer.

A final required safety measure is that before invocation of the adversary, we
need to clear (i.e. overwrite with zeros) the part of the stack that we are not
using. Otherwise, that part of the stack may still contain data and capabilities
from previously executed code, including local capabilities since they are
allowed on the stack. The adversary could take advantage of this by using it to
store either stack pointers or return pointers from previous invocations and
play the same tricks as above. Clearing the entire unused part of the stack is
the most problematic aspect of our discipline in terms of efficiency, since it
may consist of a large part of memory. We believe that capability machines like
CHERI should consider to provide special support for this requirement, by
implementing an optimized instruction for clearing (i.e. overwriting with zeros)
a large block of memory. We have no experience in processor design, but we
imagine this might be optimized as a vector instruction or by having the
processor keep track of a range of memory for which loads should return zero
irrespective of the actual contents of the memory.

\textbf{Returning from the adversary} So return pointers must be passed as local
capabilities. But what should their permissions be, what memory should they
point to and what should that memory (the activation record) contain? Let us
answer the last question first by considering what should happen when the
adversary jumps to a return pointer and what capabilities we need for achieving
that. In that case, the program counter should be restored to point after the
instruction that jumped into the adversary, so the activation record should
store this old program counter. Additionally, the stack pointer should also be
restored to its original value. Since the adversary has a more restricted
authority over the stack than the code making the call, we cannot hope to
reconstruct the return pointer from the stack pointer owned by the adversary.
Instead, the old stack pointer should also be stored as part of the activation
record.

@@

It's important that neither of these capabilities are accessible to the
adversary. In other words, the return pointer provided to the adversary must be
a capability that he/she can jump to but not read from, i.e. an enter
capability. To make this work, we construct the activation record as depicted in
Figure~\ref{fig:activ-rec-struct}. The return pointer is an $\entry$ capability
with authority over the activation record containing the previous return and
stack pointer as well as a number of restore instructions. The return pointer's
current address points to those instructions, so that, upon invocation, they are
executed and can load the old stack pointer and program counter back into the
register file. Because the return pointer is given to the adversary as an enter
pointer, he/she cannot get hold of the contents of the data structure, but after
invocation, its permission is updated to $\exec$, so the activation record's
contents become available to the restore instructions.

The final question that remains is: where should we store this activation
record? The attentive reader may already see that there is only one possibility:
since the activation record contains the old stack pointer, which is necessarily
local, the activation record can only be constructed in a part of memory where
we have write-local access, i.e. on the stack. Note that this means we will be
placing and executing instructions on the stack, i.e. it will not just contain
code pointers and data. This means that our discipline should be combined with
protection against stack smashing attacks (i.e. buffer overflows on the stack
overwriting activation records' contents), but conveniently, the fine-grained
memory protection on capability machines is well-suited to easily provide such
protection.

\textbf{Invoking an adversary callback} If we have a higher-order interface to
the adversary, we might need to invoke an adversary callback. In this case, not
so much changes with respect to the situation where we invoke regular adversary
code. The adversary can provide a callback as a capability for us to jump to,
either an $\entry$-capability if he wants to protect himself from us or just an
$\exec$ capability if he/she is not worried about that. However, there is one
scenario that we need to prevent, namely that he/she constructs the callback
capability to point into the stack, where he/she may have stored local
capabilities that he should not have access to upon invocation of the callback.
Specifically, he might store return pointers and stack pointers from previous
stack frames there, and illegally use them inside the callback.

To prevent this, we only accept callbacks from the adversary in the form of
global capabilities and we check this prior to invoking them (and fail
otherwise). This should not be an overly strict requirement, since our own
callbacks do not contain local data themselves, so there should be no need for
the adversary to construct callbacks on the stack.\footnote{Note that it does
  prevent a legitimate but non-essential scenario where the adversary wants to
  give us temporary access to a callback not allocated on the stack.}

\paragraph{Having a callback invoked by the adversary}
The above leaves us with probably the hardest scenario: how to provide a
callback to the adversary. The basic idea is that we allocate a block of memory
using malloc that we fill with the capabilities and data that the callback needs
access to, as well as some prelude instructions that load the data into
registers and jumps to the right code. We can then provide the adversary with an
enter-capability covering the allocated block and pointing to the contained
prelude instructions. However, the question that remains in this setup is: where
do we get the stack pointer?

Our answer is that the adversary should provide it to us, much like we provide
him with a current stack pointer when we invoke his code.  However, it is
important that we do not just accept any capability as a stack pointer but check
that it is safe to use.  Specifically, we check that it is indeed a $\rwlx$
pointer and that its current address points just below its start address.
Without these checks, err...
\dominique{is this needed for security or just for convenience?}

TODO:
\begin{itemize}
\item Reusable macro-instructions for stack and return pointer management
\begin{itemize}
\item taking care of the stuff above
\end{itemize}
\end{itemize}
